---
title: Differential expression analysis using DEseq2
author: Package Build
date: '2023-11-01'
slug: []
categories: []
tags: []
---


```{r setup, include=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

I borrow the data from this article ( https://www.pnas.org/doi/10.1073/pnas.2008762117#data-availability) to analyze Alzheimer’s disease (AD) data, aiming to show the standard analysis during my work. The previous blog mainly focused on the quality control, dimensionality reduction, clustering, and cell type annotation of single-nuclei RNA seq data, and further analyzes the data for differential abundance analysis. Here, I apply DEseq2 for differential expression analysis to identify significant genes and explore their functions.


### Libraries
```{r,results='hide',warning=FALSE,message=FALSE}
library(Seurat)
library(scuttle)
library(DESeq2)
library(dplyr)
library(EnhancedVolcano)
library(annotables)
library(clusterProfiler)
library(org.Hs.eg.db)
```

### Data
```{r}
seurat <- readRDS("../../../../../transcriptome_protein_AD_plat/sc_RNA/Seu_PCA_UMAP_annotated.RDS")
DimPlot(seurat, reduction = "umap",label = TRUE)+NoLegend()
DimPlot(seurat, reduction = "umap", group.by = "group_id")
```

```{r, fig.width=10,fig.height=5}
DimPlot(seurat, reduction = "umap", split.by = "orig.ident")+NoLegend()
```

From above plots, we can see that cells in NC11 is much lesser than other samples. Here, I assume NC11 as a outlier.


### Creating pseudo-bulk samples

```{r}
sce <- as.SingleCellExperiment(seurat)
summed <- aggregateAcrossCells(sce, 
    ids=colData(sce)[,c("orig.ident", "cell_type")])
summed <- summed[,summed$orig.ident != "NC11"]


```



### DE analysis and visualization
```{r}
dir.create("DE_res")
ht_list <- list()
pca_list <- list()
for (i in unique(summed$cell_type)) {
  current <- summed[,summed$cell_type == i]
  ## Remove lowly expressed genes which have less than 6 cells with any counts
  current <- current[rowSums(counts(current) > 1) >= 5,]
  cluster_counts <- counts(current)
  cluster_metadata <- data.frame(colData(current))
  dds <- DESeqDataSetFromMatrix(cluster_counts, 
                              colData = cluster_metadata, 
                              design = ~ group_id)
  # Transform counts for data visualization
  # rld <- rlog(dds, blind=TRUE)
  

  # Run DESeq2 differential expression analysis
  dds <- DESeq(dds)
  compare_list <- c("AD","NC")
 
    contrast <- c("group_id", compare_list)
    res.all <- results(dds, contrast = contrast) %>% data.frame()
    write.csv(res.all, paste0("DE_res/",i,".csv"))
    
    res.other <- res.all[res.all$pvalue >= 0.001,]
    
  # Here select 20% insignificant genes randomly for plot. 
    set.seed(123)
    res.sel <- res.other[sample(1:nrow(res.other), nrow(res.other) * 0.2),]
    res <- rbind(res.all[res.all$pvalue < 0.001,], res.sel)
    ## volcanoplot
    p <- EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(min(res$log2FoldChange, na.rm = TRUE) - 1,max(res$log2FoldChange, na.rm = TRUE) + 1),
    ylim = c(0, max(-log10(res$pvalue), na.rm = TRUE) + 1),
    title = i,
    pCutoff = 0.001,
    FCcutoff = 1,
    pointSize = 1.4,
    labSize = 2.3,
    axisLabSize = 8,
    titleLabSize = 10,
    subtitleLabSize = 8,
    captionLabSize = 8,
    legendLabSize = 7,
    legendIconSize = 2,
    # max.overlaps = Inf,
    drawConnectors = TRUE,
    widthConnectors = 0.3)
    
    print(p)
  
}
```

Volcano plots in each cell type. Here cutoff for the significant genes is: P value < 0.001, log2FC > 1.


### Functional analysis

Here, I explore the significant genes in "Interneuron; GABAergic neuron"  cell type. Because the  previous post in https://xiwen514.github.io/post/2023-10-30-diff-abun-analysis/ shows that the p value of this cell type is the lowerst, 0.1.

```{r}
head(grch37)
res.all <- read.csv("DE_res/Interneuron; GABAergic neuron.csv")
colnames(res.all)[1] <- "gene"
res_tableOE_tb <- res.all 


idx <- grch37$symbol%in% res_tableOE_tb$gene
ids <- grch37[idx, ]

## The gene names can map to more than one Ensembl ID (some genes change ID over time), 
## so we need to remove duplicate IDs prior to assessing enriched GO terms
non_duplicates <- which(duplicated(ids$symbol) == FALSE)
ids <- ids[non_duplicates, ] 

## Merge the IDs with the results 
res_ids <- inner_join(res_tableOE_tb, ids, by=c("gene"="symbol"))   


## Create background dataset for hypergeometric testing using all genes tested for significance in the results
allOE_genes <- as.character(res_ids$gene)
## Extract significant results
sigOE <- res_ids %>% dplyr::filter(padj< 0.05 & abs(log2FoldChange) > 1)
sigOE.order<- sigOE[order(sigOE$pvalue,decreasing = F),]
head(sigOE.order)


```


```{r,fig.width=7,fig.height=7}
## Run KEGG enrichment analysis 
ekegg <- enrichKEGG(gene = sigOE$entrez, 
                organism = "hsa",
                keyType = "kegg",
                pvalueCutoff = 1,
                pAdjustMethod = "fdr",
                universe = res_ids$entrez,
                minGSSize = 10,
                maxGSSize = 500,
                qvalueCutoff = 1,
                use_internal_data = FALSE)
                

## Dotplot 
dotplot(ekegg, color = "pvalue",showCategory=20)
```


```{r,fig.width=7,fig.height=9}
## Run GO enrichment analysis 
ego <- enrichGO(gene = sigOE$gene, 
                universe = allOE_genes,
                keyType = "SYMBOL",
                OrgDb = org.Hs.eg.db, 
                ont = "BP", 
                pAdjustMethod = "fdr", 
                pvalueCutoff = 1, 
                qvalueCutoff = 1,
                readable = TRUE)
                

## Dotplot 
dotplot(ego, color = "pvalue",showCategory=20)
```
